### 2014-09-23 RFCs/HDF5/VOL/developer_guide/body.tex
more updates
Former-commit-id: 71343f270c3f1a4a23269c10a4fd82cf50bcc727

\section{The Named Datatype Pickle}

\section{Dynamically Loaded VOL Plugins}

\section{Adding New Features to the Library}

>>>>

\section{The Named Datatype Pickle}
Datatypes in HDF5 exist in two states, transient and committed. When a user creates an HDF5 datatype, it is in the transient state, i.e. in memory space. It does not exist in the file; in fact, an HDF5 program could work with datatypes without even opening or accessing any HDF5 file. HDF5 provides a way for users to write (or commit) a datatype to the HDF5 file. When a user calls {\tt H5Tcommit()} on a transient type, that datatype is written to the HDF5 file under the location specified in the commit call. The datatype then becomes a named/committed datatype. Since the datatype is not stored in the file, the use can open that datatype from the file later to retrieve it and work with it in a similar way like other objects that are stored in the file (groups, datasets, and attributes).

There are a large set of routines that operate on datatypes in general. For a complete list, look at the reference manual for the {\tt H5T} interface. As far as the user is concerned, it does not matter whether the datatype is committed or not. Those routines work the same way for transient or named datatypes. With the addition of the VOL, a datatype class is added to handle committing and opening a named datatype, in addition to retrieving some metadata stored with the named datatype. However other operations that work with datatypes in both its transient or committed state do not pass through the VOL, since they do not access the datatype in the container. At that point the datatype should already by open and loaded into memory space for those operations to work. We chose not to handle those operations through the VOL even if that datatype is committed, to relieve VOL plugin developers from implementing a large set of datatype routines that is already available in the HDF5 library and is likely to be the same across all VOL plugins.

The design decision above does have some implications on the internal handling of named datatypes with the VOL changes. The higher level datatype APIs expect that the identifier for dataypes (named and transient) to map to the private {\tt H5T\_t} structure. However; as we explained earlier in this document, the object that we get back from the VOL plugins when committing a datatype to the file is a {\tt void} pointer that the library just needs to keep track. For the native plugin we do know that the pointer is actually pointing to the internal {\tt H5T\_t} structure for the name datatype, but we can not make that assumption for other plugins that will return a pointer to their own implementation of named datatypes. Furthermore we need to wrap whatever object returned from the plugin when creating/committing a named datatype with the higher level object structure that contains the VOL plugin information used for creating the datatype. This mismatch between what the VOL expects and what the {\tt H5T} API functions expect will require us to do some special handling for committing and opening named datatypes:

\begin{itemize}
\item Committing a transient datatype: At this point the identifier for the datatype exists and points to the private transient {\tt H5T\_t} structure. The difference between other objects and named datatypes is that the identifier for the named datatype should always dereference to the private {\tt H5T\_t} structure, so instead of updating the identifier's underlying object after committing the datatype to the newly created {\tt H5VL\_object\_t} structure for the VOL object of the named datatype, we store that object pointer in a newly added field to the private {\tt H5T\_t} structure and use that field to determine if that datatype is committed or not. At this point we satisfied the requirement for the public {\tt H5T} functions that do not care about the state of the datatype and can continue to work with the datatype identifier since it still dereferences to the same {\tt H5T\_t } structure, but with an additional field. The additional field contains the VOL object that should be used instead of the {\tt H5T\_t} structure when operating on that datatype through the VOL.

\item Opening a named datatype: The open operation for a named datatype is forwarded to the VOL that returns the VOL specific pointer to the datatype that is opened. The open operation for datatypes should return an identifier that dereferences to an {\tt H5T\_t} structure for the reasons we stated earlier. However at this point we do not have the {\tt H5T\_t} structure for that datatype. For this reason we require all VOL plugins that support named datatypes to store a serialized copy of the HDF5 datatype somewhere in the underlying container when committing a named datatype. After opening the datatype through the VOL, we ask the VOL plugin to return the serialized form of the datatype which we can call {\tt H5T\_decode()} on to create the private {H5T\_t} structure which is a transient form of the datatype. Then we store the object created from the open operation in the new {\tt vol\_obj} field of the {\tt H5T\_t} structure we just created and atomize it to return to the user.
\end{itemize}

HDF5 datatypes proved to be quite a challenge with the VOL due to the two states that they exist in unlike any other HDF5 object. However we believe the design decisions we made and explained above are a compromise to ease the work for plugin developers and HDF5 library maintainers at the same time.

%\section{Dynamically Loaded VOL Plugins}
%The VOL supports loading third party VOL plugins that have been installed on a system as shared libraries. The implementation for checking and loading those plugins is similar to the HDF5 filter plugin loading.

\section{Adding New Features to the Library}
It is expected that new features and routines will be added to the HDF5 library after the VOL has been introduced. New functionality added that accesses data in the file has to go through the VOL the same way that any other function does. There are several ways to handle extensions to the VOL.

\begin{itemize}
\item Adding a new VOL class for the new feature. This usually means that a new object type is added to the HDF5 library or a new service is introduced that is expected to be general in the sense that most plugins should be implemented. For example, consider the case when a new HDF5 object is added to the library that allows users to create and use a KV store in the HDF5 container. Adding a new class means that all existing plugins have to be modified to cope with the change in the VOL public structure. Changing the VOL public structure should be done only in major HDF5 releases, and not in maintenance releases.

\item Adding a new callback to a VOL class. This is done when a new API function is added to a particular class of operations (Files, Datasets, Groups, etc...) and is quite important that it can't be ``hidden'' in the {\tt specific} callback of that class. This change in the public VOL class also requires all existing plugins to be updated and should be done in major releases only.

\item Adding a new operation type to the {\tt get, specific} or {\tt optional} callbacks of a sub-class in the VOL public structure. When a new operation type to retrieve data from a particular object or to execute some new operation on that object, whether that operation is general among all plugins or plugin specific, adding a new type to the {\tt enum} of types for the three callbacks present in every sub-class should be enough to support the new operation. For more information on how this works, please consult with the VOL user guide.

\item Adding a new operation type to the {\tt optional} callback in the VOL public class for services. New HDF5 functionality that is added specifically for certain plugins and would not make sense to promote to an actual class since it isn't general enough to be implemented by most plugins or would not be used a lot by applications should follow this option.

\end{itemize}
